A new tidy data structure to support exploration and modeling of temporal data

### 2019-08-21 tsibble.Rmd
done with conclusion revision

Despite this variety and heterogeneity of temporal data, the lack of a uniform data organization expects analysts to do their own data preprocessing and take care of anything else needed to allow further analysis, which leads to a myriad of ad hoc solutions and duplicated efforts.

@r4ds proposed the tidy data workflow, which provides a conceptual framework for exploring data (as described in Figure \@ref(fig:tsibble-fit)). In temporal domain, data with temporal information arrives at the "import" stage. A new abstraction, *tsibble*, is the gatekeeper at the "tidy" stage, to verify if the raw temporal data is appropriate for downstream analytics. The exploration loop will be aided with declarative grammars for more robust and accurate analysis.

@@

Time series can be univariate or multivariate, and for modeling require relatively long lengths (i.e., large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. Most statistical software represent such time series as vectors or matrices. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wider format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric vectors. Time indices are implicitly inferred as attributes or meta-information. The number of observations must be the same across all the series. Data wrangling, from the form that data arrives in to this specialist format, can be mysterious and frustrating, inhibiting the robustness of downstream tasks. This kind of format is model-centric, and the diagram in the style of Figure \@ref(fig:tsibble-fit) would place the "model" at the center of the analytical universe, and all the transformations and visualizations would hinge on that format. This is contrary to the **tidyverse** conceptualization, which holistically captures the full data workflow.

@@

Stata [@Stata] provides two commands, `tsset` and `xtset`, to declare time series and panel data respectively, both of which need specifying panel and time variables. Note they are in fact the same command, according to the manual. The underlying data arrangement is the same and of longer form. Either time series or panel commands are sufficient to apply to both types. The SAS software [@SAS-STAT] handles them in a similar way as Stata.

@@

The new tsibble structure bridges the gap between wild temporal data and the tamed state of model inputs. The proposed contextual semantics is the new addition to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series and panel model inputs. Index, key and time interval build the three stone pillars to this new semantically structured temporal data. Each is now described in more detail.

@@

Most time series models are univariate, such as ARIMA and Exponential Smoothing, modelling temporal dynamics for each series independently. The **fable** package [@R-fable] provides a tidy forecasting framework built on top of tsibble is under development, with the goal of promoting transparent and human-centered forecasting practices. With the presence of the key, tsibble holds many series. Since models are fundamentally scalable, the `model()` and `forecast()` generics will take care of fitting and forecasting univariate models to each series across time in a tsibble at once.

@@

The "tsibble" is a new data abstraction to represent temporal data, allowing the "tidy data" principles to be brought to the time domain. Tidy data begins to take shape in the state of time with the introduction of the contextual semantics of index and key. A declared index provides direct support to the time variable; variables that comprise the key define observations over time. These semantics further determine unique data entries required for a valid tsibble. No matter how temporal data arrives, a tsibble respects a time index and maintains the data richness. A tsibble frictionlessly allows transformation, visualization and modeling, and smoothly shifts between them, allowing for rapid iteration to gain data insights.

* extensive index
* flexible key

A missing piece of the **tsibble** package is to enable user-defined calendars and to respect structurally missing observations. For example, a call center may operate only between 9:00 am and 5:00 pm on week days, and stock trading resumes on Monday straight after Friday. No data available outside trading hours would be labeled as structural missingness, which tsibble currently disregards. However, a few R packages provide functionality to create and manage many sorts of calendars, including market-specific business calendars. <!-- This delays the implementation.--> Generally, custom calendars are easily embedded into the tsibble framework. Consequently these tsibble operators, like `fill_gaps()`, would work out of the box, and forecasts would be generated within its definable time range.

The **tsibble** package provides the grammar of temporal data manipulation, regardless of how the data is stored. Currently, it works for managing and manipulating temporal data frames in memory locally. But it is possible to work with remote tables stored in databases, such as SQLite and MySQL, using exactly the same tsibble code. This is left for future work.

@@

The authors would like to thank Mitchell O'Hara-Wild for many discussions on the software development and Davis Vaughan for contributing ideas on rolling window functions. We also thank Stuart Lee for the feedback on this manuscript. This article was created with knitr [@knitr] and R Markdown [@rmarkdown]. The project's Github repository <https://github.com/earowang/paper-tsibble> houses all materials required to reproduce this article and a history of the changes.

```{r write-bib}
write_bib(c("dplyr", "tidyr", "purrr", "nanotime", "tibbletime", "tsibble", "magrittr", "future", "furrr", "bomrang", "rlang", "fable", "panelr"), file = "rpkgs.bib")
```

>>>>

Despite this variety and heterogeneity, little organization or conceptual oversight on how one should get the wild data into a tamed state is provided for temporal data. Analysts are expected to do their own data preprocessing and take care of anything else needed to allow further analysis, which leads to a myriad of ad hoc solutions and duplicated efforts.

@r4ds proposed the tidy data workflow, which provides a conceptual framework for exploring data (as described in Figure \@ref(fig:tsibble-fit)). In temporal domain, data with time information arrives at the "import" stage. A new abstraction, *tsibble*, is the gatekeeper at the "tidy" stage, to verify if the raw temporal data is appropriate for downstream analytics. The exploration loop will be aided with declarative grammars, yielding more robust and accurate analyses.

@@

Time series can be univariate or multivariate, and for modeling require relatively long lengths (i.e., large $T$). Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic processes, for the primary purpose of forecasting, and characterizing temporal dynamics. Most statistical software represent such time series as vectors or matrices. Multivariate time series are typically assumed to be in the format where each row is assumed to hold observations at a time point and each column to contain a single time series. (The tidy data name for this would be **wider format**.) This implies that data are columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric vectors. Time indices are implicitly inferred as attributes or meta-information. The number of observations must be the same across all the series. Data wrangling, from the form that data arrives in to this specialist format, can be mysterious and frustrating. This kind of format is model-centric, and the diagram in the style of Figure \@ref(fig:tsibble-fit) would place the "model" at the center of the analytical universe, and all the transformations and visualizations would hinge on that format. This is contrary to the **tidyverse** conceptualization, which holistically captures the full data workflow.

@@

Stata [@Stata] provides two commands, `tsset` and `xtset`, to declare time series and panel data respectively, both of which need specifying panel and time variables. Note they are in fact the same command, according to the manual. The underlying data arrangement is the unifying longer form. Either time series or panel commands are sufficient to apply to both types. The SAS software [@SAS-STAT] handles them in a similar way as Stata.

@@

The new tsibble structure bridges the gap between raw data and the rigorous state of temporal data analysis. The proposed contextual semantics is the new add-on to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series and panel model inputs. Index, key and time interval build the three stone pillars to this new semantically structured temporal data. Each is now described in more detail.

@@

Most time series models are univariate, such as ARIMA and Exponential Smoothing, modelling temporal dynamics for each series independently. The **fable** package [@R-fable] provides a tidy forecasting framework built on top of tsibble, which is under development, with the goal of promoting transparent and human-centered forecasting practices. With the presence of the key, tsibble holds many series. Since models are fundamentally scalable, the `model()` and `forecast()` generics will take care of fitting and forecasting univariate models to each series across time in a tsibble at once.

@@

A data abstraction, tsibble, is proposed to represent temporal data, allowing the "tidy data" principles to be brought to the time domain. Tidy data begins to take shape in the state of time with new contextual semantics: index and key. The index variable provides direct support to an exhaustive set of ordered objects. The key identifies observational units over time, which can consist of single or multiple variables to be useful in more scenarios. These semantics further determine unique data entries required for a valid tsibble. It shepherds raw temporal data through the "tidy" stage to the next exploration stage for gaining more insights.

The supporting toolkits articulate the temporal data pipeline, with the shared goal of reducing the time between the framings of data questions and the code realization. The rapid iteration for better understanding data is achieved through frictionlessly shifting among transformation, visualization, and modelling, using the standardized tsibble data infrastructure.

A nice feature that the tsibble structure should take into account is to allow user-defined calendars and to respect structurally missing observations. For example, a call center may operate only between 9:00 am and 5:00 pm on week days, and stock trading resumes on Monday straight after Friday. No data available outside trading hours would be labeled as structural missingness, which tsibble currently disregards. Customer calendars can be embedded into the tsibble framework in theory. A few R packages provide functionality to create and manage many specific calendars, such as the **bizdays** package [@R-bizdays] for business days calendars. However, the lack of a generic calendar system delays the implementation. This is left for future work.

@@

The authors would like to thank Mitchell O'Hara-Wild for many discussions on the software development and Davis Vaughan for contributing ideas on rolling window functions. We also thank Stuart Lee for the feedback on this manuscript. We are grateful for anonymous reviewers for helpful feedback that has led to many improvements of the paper. This article was created with knitr [@knitr] and R Markdown [@rmarkdown]. The project's Github repository <https://github.com/earowang/paper-tsibble> houses all materials required to reproduce this article and a history of the changes.

```{r write-bib, eval = FALSE}
write_bib(c("dplyr", "tidyr", "purrr", "nanotime", "tibbletime", "tsibble", "magrittr", "future", "furrr", "bomrang", "rlang", "fable", "panelr", "bizdays"), file = "rpkgs.bib")
```
